Models

### 2023-02-14 Code/Models.Rmd
Fixed clusrank

##Models
```{r}
men_finals_yeareffect <- lmer(RxnTime ~ (1|Year), data = men_finals)
men_pooled_yeareffect <- lmer(RxnTime ~ (1|Year), data = men_pooled)
old_finals_yeareffect <- lmer(RxnTime ~ (1|Year), data = old_finals)
old_pooled_yeareffect <- lmer(RxnTime ~ (1|Year), data = old_pooled)
AIC(men_finals_yeareffect)
AIC(old_finals_yeareffect)
AIC(men_pooled_yeareffect)
AIC(old_pooled_yeareffect)
```

```{}
Thus this seems to support that the 2022 data should be included and that the semifinals and finals should be pooled together.
```

```{r}
trial <- subset(times, Gender == "M" & Time != 'DQ')
x <- lmer(RxnTime ~ (1|Year), data = trial)
AIC(x)
```

```{}
This shows that increasing the data set increases AIC, unsure what we want to use
```

```{r}
#men_finals_randeffect <- lmer(RxnTime ~ 1, data = men_pooled)
men_pooled_yeareffect <- lmer(RxnTime ~ (1|Year), data = men_pooled)
men_pooled_randeffect <- glm(RxnTime ~ 1, data = men_pooled)

anova(men_pooled_yeareffect, men_pooled_randeffect)

```
```{}
As the year effect model is signficant at all alpha levels and the AIC is lower, the year effect is significant and should be considered in the model.  Now we compare the log model to the year effect model
```

```{r}
men_pooled_yeareffect <- lmer(RxnTime ~ (1|Year), data = men_pooled)
men_pooled_log_year <- lmer(log(RxnTime) ~ (1|Year), data = men_pooled)
men_pooled_log <- glm(log(RxnTime) ~ 1, data = men_pooled)
anova(men_pooled_yeareffect, men_pooled_log_year, men_pooled_log)
```
```{}
The year effect model is better than both the log random effect model and the log year effect model.
```

```{r}
library(lme4)
men_pooled_gamma_year <- glmer(RxnTime ~ (1 | Year), data = men_pooled, family = Gamma)
men_pooled_log_gamma_year <- glmer(RxnTime ~ (1 | Year), data = men_pooled, family = Gamma(link = "log"))
men_pooled_yeareffect <- lmer(RxnTime ~ (1|Year), data = men_pooled)

#Compare how much the link=log matters
anova(men_pooled_gamma_year, men_pooled_log_gamma_year)

#Compare gamma and linear
anova(men_pooled_gamma_year, men_pooled_yeareffect)

```
```{}
Whether or not the model contains "(link = "log")" seems to have very little effect on the strength of the model.  The gamma model seems to perform better than the linear model based on AIC, but due to the gamma model being more difficult to interpret, the linear mixed effects model seems sufficient.
```

##Simulations
```{r}
summary(men_pooled_gamma_year)
qqnorm(resid(men_pooled_yeareffect))
qqline(resid(men_pooled_yeareffect))
qqGamma(resid(men_pooled_gamma_year))
```

```{r}
library(qpToolkit)
qqGamma(resid(men_pooled_gamma_year))
```

```{r}
library(DHARMa)
men_pooled_gamma_year <- glmer(RxnTime ~ (1 | Year), data = men_pooled, family = Gamma(link = "log"))
summary(men_pooled_gamma_year)
```
```{r}
simout_gamma <- simulateResiduals(men_pooled_gamma_year, n = 1000)
simout_linear <- simulateResiduals(men_pooled_yeareffect, n = 1000)
plot(simout_gamma)
plot(simout_linear)
#summary(simout)
#str(simout)

```

##2019 vs 2022 Comparisons
```{r}
comp <- read.csv("../Data/2019vs2022.csv")
#"C:\Users\ofior\Documents\Thesis\Data\2019vs2022.csv"
colnames(comp) <- c("Athlete", "Country", "NineteenH", "NineteenS", "NineteenF", "TwentyTwoH", "TwentyTwoS", "TwentyTwoF")
#View(comp)
```

```{r}
library(clusrank)
t.test(comp$NineteenH, comp$TwentyTwoH, paired = TRUE)
clusWilcox.test(comp$NineteenH, cluster=comp$TwentyTwoH, paired = TRUE, alternative = "greater")
wilcox.test(comp$NineteenH, comp$TwentyTwoH, paired = TRUE)
```

```{r}
t.test(comp$NineteenS, comp$TwentyTwoS, paired = TRUE)
clusWilcox.test(comp$NineteenS, cluster=comp$TwentyTwoS, paired = TRUE, alternative = "greater")
wilcox.test(comp$NineteenS, comp$TwentyTwoS, paired = TRUE)
```

```{}
A matched pairs t test, clusrank package Wilcox test, and a standard Wilcox test all showed that the heat means differ by a significant amount. And two of the three semifinals means had an alpha less than 0.01.
```

```{r}
nations <- read.csv("../Data/NatVSWorld.csv")
colnames(nations) <- c("Athlete", "Country", "NatH", "NatS", "NatF", "WorldH", "WorldS", "WorldF")
```

```{r}
library(clusrank)
t.test(nations$NatH, nations$WorldH, paired = TRUE)
clusWilcox.test(nations$NatH, cluster=nations$WorldH, paired = TRUE, alternative = "greater")
wilcox.test(nations$NatH, nations$WorldH, paired = TRUE)
```
```{}
These p values all significant at alpha = 0.05, but not at alpha = 0.01
```

```{r}
library(clusrank)
t.test(nations$NatS, nations$WorldS, paired = TRUE)
clusWilcox.test(nations$NatS, cluster=nations$WorldS, paired = TRUE, alternative = "greater")
wilcox.test(nations$NatS, nations$WorldS, paired = TRUE)
```
```{}
These p values are all significant at alpha = 0.01
```

>>>>

```{r}
library(lme4)
men_finals_one <- glm(RxnTime ~ 1, data = men_finals)
men_finals_year <- lmer(RxnTime ~ (1|Year), data = men_finals)
men_finals_one_gamma <- glm(RxnTime ~ 1, data = men_finals, family = Gamma)
men_finals_year_gamma <- glmer(RxnTime ~ (1|Year), data = men_finals, family = Gamma)
men_finals_one_gamma_log <- glm(RxnTime ~ 1, data = men_finals, family = Gamma(link = "log"))
men_finals_year_gamma_log <- glmer(RxnTime ~ (1|Year), data = men_finals, family = Gamma(link = "log"))

paste("Linear without year effect has AIC of", AIC(men_finals_one), "and log-lik of", logLik(men_finals_one))
paste("Linear with year effect has AIC of", AIC(men_finals_year), "and log-lik of", logLik(men_finals_year))
paste("Gamma without year effect has AIC of", AIC(men_finals_one_gamma), "and log-lik of", logLik(men_finals_one_gamma))
paste("Gamma with year effect has AIC of", AIC(men_finals_year_gamma), "and log-lik of", logLik(men_finals_year_gamma))
paste("Gamma and log without year effect has AIC of", AIC(men_finals_one_gamma_log), "and log-lik of", logLik(men_finals_one_gamma_log))
paste("Gamma and log with year effect has AIC of", AIC(men_finals_year_gamma_log), "and log-lik of", logLik(men_finals_year_gamma_log))
```
```{}
The best selected model is men_finals_year_gamma as that has the smallest AIC: -491 and largest log-lik of 248.5
```

```{r}
library(lme4)
old_finals_one <- glm(RxnTime ~ 1, data = old_finals)
old_finals_year <- lmer(RxnTime ~ (1|Year), data = old_finals)
old_finals_one_gamma <- glm(RxnTime ~ 1, data = old_finals, family = Gamma)
old_finals_year_gamma <- glmer(RxnTime ~ (1|Year), data = old_finals, family = Gamma)
old_finals_one_gamma_log <- glm(RxnTime ~ 1, data = old_finals, family = Gamma(link = "log"))
old_finals_year_gamma_log <- glmer(RxnTime ~ (1|Year), data = old_finals, family = Gamma(link = "log"))

paste("Linear without year effect has AIC of", AIC(old_finals_one), "and log-lik of", logLik(old_finals_one))
paste("Linear with year effect has AIC of", AIC(old_finals_year), "and log-lik of", logLik(old_finals_year))
paste("Gamma without year effect has AIC of", AIC(old_finals_one_gamma), "and log-lik of", logLik(old_finals_one_gamma))
paste("Gamma with year effect has AIC of", AIC(old_finals_year_gamma), "and log-lik of", logLik(old_finals_year_gamma))
paste("Gamma and log without year effect has AIC of", AIC(old_finals_one_gamma_log), "and log-lik of", logLik(old_finals_one_gamma_log))
paste("Gamma and log with year effect has AIC of", AIC(old_finals_year_gamma_log), "and log-lik of", logLik(old_finals_year_gamma_log))
```
```{}
The best selected model is old_finals_year_gamma as that has the smallest AIC: -465 and largest log-lik of 235.5.  The models for the older data does not seem to fit the data as good as for the full data set.
```

```{r}
library(lme4)
men_pooled_one <- glm(RxnTime ~ 1, data = men_pooled)
men_pooled_year <- lmer(RxnTime ~ (1|Year), data = men_pooled)
men_pooled_one_gamma <- glm(RxnTime ~ 1, data = men_pooled, family = Gamma)
men_pooled_year_gamma <- glmer(RxnTime ~ (1|Year), data = men_pooled, family = Gamma)
men_pooled_one_gamma_log <- glm(RxnTime ~ 1, data = men_pooled, family = Gamma(link = "log"))
men_pooled_year_gamma_log <- glmer(RxnTime ~ (1|Year), data = men_pooled, family = Gamma(link = "log"))

paste("Linear without year effect has AIC of", AIC(men_pooled_one), "and log-lik of", logLik(men_pooled_one))
paste("Linear with year effect has AIC of", AIC(men_pooled_year), "and log-lik of", logLik(men_pooled_year))
paste("Gamma without year effect has AIC of", AIC(men_pooled_one_gamma), "and log-lik of", logLik(men_pooled_one_gamma))
paste("Gamma with year effect has AIC of", AIC(men_pooled_year_gamma), "and log-lik of", logLik(men_pooled_year_gamma))
paste("Gamma and log without year effect has AIC of", AIC(men_pooled_one_gamma_log), "and log-lik of", logLik(men_pooled_one_gamma_log))
paste("Gamma and log with year effect has AIC of", AIC(men_pooled_year_gamma_log), "and log-lik of", logLik(men_pooled_year_gamma_log))
```
```{}
The best selected model is men_pooled_year_gamma_log as it had the smallest AIC and the largest log-lik
```

```{r}
library(lme4)
old_pooled_one <- glm(RxnTime ~ 1, data = old_pooled)
old_pooled_year <- lmer(RxnTime ~ (1|Year), data = old_pooled)
old_pooled_one_gamma <- glm(RxnTime ~ 1, data = old_pooled, family = Gamma)
old_pooled_year_gamma <- glmer(RxnTime ~ (1|Year), data = old_pooled, family = Gamma)
old_pooled_one_gamma_log <- glm(RxnTime ~ 1, data = old_pooled, family = Gamma(link = "log"))
old_pooled_year_gamma_log <- glmer(RxnTime ~ (1|Year), data = old_pooled, family = Gamma(link = "log"))

paste("Linear without year effect has AIC of", AIC(old_pooled_one), "and log-lik of", logLik(old_pooled_one))
paste("Linear with year effect has AIC of", AIC(old_pooled_year), "and log-lik of", logLik(old_pooled_year))
paste("Gamma without year effect has AIC of", AIC(old_pooled_one_gamma), "and log-lik of", logLik(old_pooled_one_gamma))
paste("Gamma with year effect has AIC of", AIC(old_pooled_year_gamma), "and log-lik of", logLik(old_pooled_year_gamma))
paste("Gamma and log without year effect has AIC of", AIC(old_pooled_one_gamma_log), "and log-lik of", logLik(old_pooled_one_gamma_log))
paste("Gamma and log with year effect has AIC of", AIC(old_pooled_year_gamma_log), "and log-lik of", logLik(old_pooled_year_gamma_log))
```
```{}
The best model is old_pooled_year_gamma as that has the smallest AIC: -1687.5 and the largest log-lik: 846.8
```

```{}
Simulate to see what prob of a time less than 0.1 second rxn time.
```

##Simulations

```{r}
library(lme4)
men_pooled_gamma_year <- glmer(RxnTime ~ (1 | Year), data = men_pooled, family = Gamma(link = "log"))
baseline <- glm(RxnTime ~ 1, data = men_pooled, family = Gamma(link = "log"))

simfit <- function(men_pooled_gamma_year, n = 12000000) {
  vc <- as.data.frame(VarCorr(men_pooled_gamma_year))
  ## mu <- exp(coef(men_pooled_gamma_year)$Year[,]) ## using fitted random effects only
  z <- rnorm(n, sd = vc$sdcor[1]) #$ using generated random effects
  mu <- exp(fixef(men_pooled_gamma_year) + z)
  shape <- 1 / vc$vcov[2]
  x <- rgamma(n, shape = shape, scale = mu / shape)
  x
}
sbaseline <- summary(baseline)
print(sbaseline)
x <- simfit(men_pooled_gamma_year)
mean(x < 0.1)
quantile(x, prob = c(.00001, .0001, .0005, .001))
## compare with and without random effect
plot(density(x)) # with random effect
curve(dgamma(x, shape = 1 / sbaseline$dispersion,
             scale = exp(coef(baseline)) * sbaseline$dispersion),
      0.08, 0.22, add = TRUE, col = "blue") # fixed effect only
```

```{r}
simulation <- function(model, base){
  simfit <- function(model, n = 12000000) {
    vc <- as.data.frame(VarCorr(model))
    ## mu <- exp(coef(model)$Year[,]) ## using fitted random effects only
    z <- rnorm(n, sd = vc$sdcor[1]) #$ using generated random effects
    mu <- exp(fixef(model) + z)
    shape <- 1 / vc$vcov[2]
    x <- rgamma(n, shape = shape, scale = mu / shape)
    x
  }
  sbase <- summary(base)
  #print(sbase)
  x <- simfit(model)
  print(mean(x < 0.1))
  print(quantile(x, prob = c(.00001, .0001, .0005, .001)))
  ## compare with and without random effect
  plot(density(x)) # with random effect
  curve(dgamma(x, shape = 1 / sbase$dispersion, 
               scale = exp(coef(base)) * sbase$dispersion), 
        0.08, 0.22, add = TRUE, col = "blue") # fixed effect only
}
simulation(old_pooled_year_gamma_log, old_pooled_one_gamma_log)
```

```{}
Every 1000 racers, we expect 7 to have a reaction time below 0.1 seconds.  If we want 1 out of every 1000
```

##2019 vs 2022 Comparisons
```{r}
comp <- read.csv("../Data/Clusrank2019vs2022.csv")
#"C:\Users\ofior\Documents\Thesis\Data\2019vs2022.csv"
colnames(comp) <- c("Time", "Group", "Cluster")
#View(comp)
```

```{r}
library(clusrank)
#t.test(comp$NineteenH, comp$TwentyTwoH, paired = TRUE)
clusWilcox.test(comp$Time, group=comp$Group, cluster=comp$Cluster, paired = TRUE, alternative = "greater", method = "ds")
clusWilcox.test(comp$Time, group=comp$Cluster, cluster = comp$Group, paired = TRUE, alternative = "greater", method = "ds")
#wilcox.test(comp$NineteenH, comp$TwentyTwoH, paired = TRUE)
```

```{}
Cluster by 2019 and 2022
Columns: Athlete, RxnTime, 2019/2022
```

```{r}
nations <- read.csv("../Data/ClusrankNatvsWorld.csv")
colnames(nations) <- c("Time", "Group", "Cluster")
```

```{r}
library(clusrank)
#t.test(nations$NatH, nations$WorldH, paired = TRUE)
clusWilcox.test(nations$Time, group=nations$Group, cluster=nations$Cluster,
                paired = TRUE, alternative = "greater", method = "ds")
clusWilcox.test(nations$Time, group=nations$Cluster, cluster = nations$Group, 
                paired = TRUE, alternative = "greater", method = "ds")
#wilcox.test(nations$NatH, nations$WorldH, paired = TRUE)
```
